---
title: "VBZ Delays Analysis"
author: "Mirai Solutions"
date: "`r Sys.time()`"
output:
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
library(dplyr)
library(vroom)
library(ggplot2)
# set to FALSE if data have been already processed
process_data <- TRUE
```

## Introduction

References:

* Data: https://data.stadt-zuerich.ch/dataset/vbz_fahrzeiten_ogd
* Names translation: https://github.com/OpenDataDayZurich2016/translations/blob/master/Translation_Data_Attributes.pdf


## Data handling

Weekly data and corresponding resource IDs:

```{r, eval=process_data}
base_url <- "https://data.stadt-zuerich.ch/dataset/vbz_fahrzeiten_ogd/resource"
vbz_resources <- c(
  "20190714_20190720" = "9beb05c6-0153-4ffa-a928-8ff899bf5f1c",
  "20190721_20190727" = "7883fd6e-e8f3-4ce4-82df-4944baa01a5b",
  "20190728_20190803" = "5bec146e-ba16-4a6a-8f6b-44e4ed14bdb9",
  "20190804_20190810" = "afdb312d-2855-4437-a8cd-f1c28dd97a79"
)
```


### Download data

Data are downloaded if not already available

```{r, eval=process_data}
raw_data_dir <- "data-raw"
dir.create(raw_data_dir)
lapply(names(vbz_resources), function(x) {
  data_file <- sprintf("fahrzeiten_soll_ist_%s.csv", x)
  path <- file.path("data-raw", data_file)
  if (file.exists(path)) {
    message("skip download of ", path)
  } else {
    download.file(
      paste(base_url, vbz_resources[[x]], "download", data_file, sep = "/"),
      destfile = path
    )
    message("downloaded ", path)
  }
}) %>% invisible()
```


### Read all data

Load and transform data with package [**vroom**](https://vroom.r-lib.org/).

```{r, eval=process_data}
files <- file.path(
  raw_data_dir, 
  sprintf("fahrzeiten_soll_ist_%s.csv", names(vbz_resources))            
)
cols_types <- vroom::cols(
  betriebsdatum = vroom::col_date("%d.%m.%y"),
  linie = "i" # integer
)
data <- vroom::vroom(
  files,
  col_types = cols_types, altrep_opts = FALSE,
  col_select = c(
    date = betriebsdatum,
    line = linie,
    scheduled_departure = dplyr::matches("soll_ab_von"),
    actual_departure = dplyr::matches("ist_ab_von"),
    scheduled_arrival = dplyr::matches("soll_an_nach"),
    actual_arrival = dplyr::matches("ist_an_nach")
  )
)
```


### Line data manipulation

* Keep only tram lines (number < 30).
* Extract the weekday.
* Define the hour as the scheduled departure time in hours from the beginning of the operating day (4AM). This implies that time after midnight is assigned and hour > 24.

```{r, eval=process_data}
data <- data %>%
  filter(line < 30) %>%
  mutate(
    weekday = lubridate::wday(
      date, label = TRUE, 
      locale = "en_US.UTF-8", week_start = 1
    ),
    hour = (scheduled_departure/3600),
    hour = ifelse(hour < 4, hour + 24, hour)
  ) %>% 
  select(grep("arr|dep", names(.), invert = TRUE, value = TRUE), everything())

data
```


### Save data by line

```{r, eval=process_data}
line_data_dir <- "line-data"
dir.create(line_data_dir)
lapply(unique(data$line), function(l) {
  line_rds <- file.path(line_data_dir, sprintf("line-%d.rds", l))
  data %>%
    filter(line == l) %>%
    saveRDS(file = line_rds)
  message("saved ", line_rds)
}) %>% invisible()
```


## Analysis

### Load line data

Select the line and read departure and arrival data

```{r}
line <- 11
line_file <- file.path("line-data", sprintf("line-%d.rds", line))
data <- readRDS(line_file)
```

### Analysis of delays for line `r line`


#### Delays by hour for each weekday

Compute delays in minutes as the mean of departure and arrival delays, and compute delay counts by hour for each weekday

```{r}
delays <- data %>%
  mutate(
    delay = ((actual_departure - scheduled_departure) +
               (actual_arrival - scheduled_arrival)) / 2 / 60
  ) %>% 
  group_by(
    weekday,
    hour = cut(hour, 0:48, labels = ifelse(0:47 < 24, 0:47, paste0("+", floor(0:47) - 24))),
    delay = cut(delay, c(-Inf, 1:5, Inf))
  ) %>% 
  summarize(count = n())

delays
```

#### Visualize results

Barplot of normalized delay counts by hour for each weekday, conditional on delays of at least 1 minute

```{r delays-plot, fig.asp=1.5}
delays %>% 
  filter(delay != "(-Inf,1]") %>%
  ggplot(aes(x = hour, y = count)) +
  geom_col(aes(fill = delay), position = "fill") +
  lemon::facet_rep_wrap(~weekday, ncol = 1, repeat.tick.labels = TRUE) +
  theme(legend.justification = "top")
```
